Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-19 paper.tex
Discuss mutations.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations.
Through the use of efficient data structures, it is
now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} or
VCF~\citep{danecek2011variant} based formats formats.
This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2019efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

@@

\label{sec-ts}
\subsection*{Tree sequences}
One of the key reasons for \msprime's significiant performance advantage
over other simulators is its usage of the ``succinct tree sequence''
(usually, tree sequence) data stucture and the supporting \tskit\
library~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we briefly summarise
here.

@@

The succinct tree sequence is a data structure for concisely encoding
genetic ancestry. Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges}. A node in tree sequence terminology corresponds
to a single chromosome strand, and is associated with its birth time,
birth population and other metadata. A node may optionally be
associated with an \emph{individual}, so that, for example, a diploid
individual will be associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within a
specific individual, and edges define genetic inheritence
relationships between these nodes. An edge consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the example edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
is a generalisation from the original concept  of
``coalescence records''~\citep{kelleher2016efficient}, and
is the key to the efficiency of tree sequences.
\jkcomment{Summarise and add references?}
As discussed in the \ref{sec-mutations} section, tree sequences
can also store genetic variation data very efficiently.

The term Ancestral Recombination Graph (ARG) is often understood
to mean the complete record of a sample's genetic
ancestry~\citep[e.g.][]{mathieson2020ancestry}, and in this sense a tree
sequence \emph{is} an ARG. More technically, an ARG is a particular
graphical representation of genetic ancestry based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information---although we do not store
the additional ARG information by default
in \msprime\ because the minimal tree sequence encoding is
much more concise (see the \nameref{sec-arg} section for more details).

The point of simulations is to generate data and the efficiency with
which simulated data can be incorporated into an overall application
is an important issue.
For example, Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and more recent machine learning
methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
perform population genetic inference by using large numbers of
simulations as training data. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence data.
Modern approaches such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large
numbers of user-specified statistics that are individually weakly informative,
but jointly close to sufficient for the desired model or parameter.
However, the computational resources needed to calculate these statistics can be even
greater than those used to generate the simulated datasets.
Older software such as DIYABC~\citep{cornuet2008inferring}
or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
developing their own specialised simulators, tightly integrating them
with the inference method. While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error.
As ABC inference has become useful in an ever-widening range of applications,
these all-in-one methods risk falling behind if they do not allow users to explore
the models and statistics of most interest.

\jkcomment{Stuff to cover}
\begin{itemize}

\item The core problem is that the field lacked an efficient mechanism for
    interchanging simulated data, and so developers were forced to choose
    between efficiency and duplication of effort.

@@

\item Side note somewhere: with efficient access to branch length
    statistics~\citep{ralph2019efficiently}, maybe we don't need to
    simulate mutations at all. (not in the ABC context, but elsewhere)

\item Msprime uses the \tskit\ library~\citep{tskit2021tskit}, which provides
    extremely efficient interchange and data processing. A quick example
    of how efficient.

@@

\item Recently, Phastsim/USHER~\cite{demaio2021phastsim,turakhia2021ultrafast}
    have used this idea of storing the
    tree plus mutations, but approach is crude compared to tskit.

@@

\label{sec-mutations}
\subsection*{Finite sites mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

>>>>

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations.
Through the use of efficient data structures, it is
now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} or
VCF~\citep{danecek2011variant} based formats formats.
This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

@@

\subsection*{Tree sequences}
\label{sec-ts}
One of the key reasons for \msprime's significiant performance advantage
over other simulators is its usage of the ``succinct tree sequence''
(usually, tree sequence) data stucture and the supporting \tskit\
library~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we briefly summarise
here.
\jkcomment{We need to mention that tree sequences and tskit are
derived from the 2016 paper and have been extracted from msprime
into a separate C/Python library.}

@@

The succinct tree sequence is a data structure for concisely encoding
genetic ancestry. Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges}. A node in tree sequence terminology corresponds
to a single chromosome strand, and is associated with its birth time,
birth population and other metadata. A node may optionally be
associated with an \emph{individual}, so that, for example, a diploid
individual will be associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within a
specific individual, and edges define genetic inheritence
relationships between these nodes. An edge consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the example edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
is a generalisation from the original concept  of
``coalescence records''~\citep{kelleher2016efficient}, and
is the key to the efficiency of tree sequences.
\jkcomment{Summarise and add references?}

The term Ancestral Recombination Graph (ARG) is often understood
to mean the complete record of a sample's genetic
ancestry~\citep[e.g.][]{mathieson2020ancestry}, and in this sense a tree
sequence \emph{is} an ARG. More technically, an ARG is a particular
graphical representation of genetic ancestry based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information---although we do not store
the additional ARG information by default
in \msprime\ because the minimal tree sequence encoding is
much more concise (see the \nameref{sec-arg} section for more details).
It should be noted that, despite efforts to
standardise~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

Tree sequences encode a complete and concise description of a set of sampled
chromosome's genetic ancestry, and they can also efficiently represent
the genetic variation (extant and ancestral).
As illustrated in Fig.~\ref{fig-ts-example}, the \tskit\ tree
sequence encoding includes \emph{site} and \emph{mutation} tables. Each
row in the site table represents some position along the genome, and
each row in the mutation table represents a change of allelic state
at a particular site and node. Since the number of mutations is typically
small, we can usually store the genetic variation for a given variant
site in constant space. This approach of storing the genetic variation
via the genetic ancestry is both very concise and extremely efficient
to process---for example, computing Tajima's $D$ using \tskit\
is several orders of magnitude faster for large sample sizes than
efficient libraries working with the variant matrix
representation~\citep{ralph2020efficiently}.
\jkcomment{Mention Phastsim/USHER~\citep{demaio2021phastsim,turakhia2021ultrafast}
here. Same basic idea, but much cruder.}

\jkcomment{Make some sort of link here between this paper, which is
about msprime simulations, but that we can't really disentangle
simulations from their downstream processing.}
The point of simulations is to generate data and the efficiency with
which simulated data can be incorporated into an overall application
is an important issue.
For example, Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and more recent machine learning
methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
perform population genetic inference by using large numbers of
simulations as training data. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence data.
Modern approaches such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large
numbers of user-specified statistics that are individually weakly informative,
but jointly close to sufficient for the desired model or parameter.
However, the computational resources needed to calculate these statistics can be even
greater than those used to generate the simulated datasets.
Older software such as DIYABC~\citep{cornuet2008inferring}
or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
developing their own specialised simulators, tightly integrating them
with the inference method. While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error.
As ABC inference has become useful in an ever-widening range of applications,
these all-in-one methods risk falling behind if they do not allow users to explore
the models and statistics of most interest.

\jkcomment{Other potential stuff to cover}
\begin{itemize}

@@

\item Side note somewhere: with efficient access to branch length
    statistics~\citep{ralph2020efficiently}, maybe we don't need to
    simulate mutations at all. (not in the ABC context, but elsewhere)

@@

@@

\subsection*{Finite sites mutations}
\label{sec-mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.
